\labeledsection{General Problem Solving}{sec:search}
\definition{Search Problem}{
A search problem $\Pi := \tuple{\mathcal{S}, \mathcal{A}, \mathcal{T}, \mathcal{I}, \mathcal{G}}$ consists of a \linkterm{set}{def:set} $\mathcal{S}$ of states, a \linkterm{set}{def:set} $\mathcal{A}$ of actions, and a transition model $\func{\mathcal{T}}{\cartprod{\mathcal{A}, \mathcal{S}}}{\powerset{\mathcal{S}}}$ that assigns to any action $a \in \mathcal{A}$ and state $s \in \mathcal{S}$ a \linkterm{set}{def:set} of successor states. Certain states in $\mathcal{S}$ are designated as goal (terminal) states ($\mathcal{G} \subseteq \mathcal{S}$ with $\mathcal{G} \neq \phi$) and initial states $\mathcal{I} \subseteq \mathcal{S}$. 
}{search_problem}

\definition{Action Application}{
We say that an action $a \in \mathcal{A}$ is applicable in state $s \in \mathcal{S}$, iff $\mathcal{T}(a,s) \neq \phi$ and that any $s' \in \mathcal{T}(a,s)$ is a result of applying action $a$ to state $s$. We call $\func{\mathcal{T}_a}{\mathcal{S}}{\powerset{\mathcal{S}}}$ with $\mathcal{T}_a(s) := \mathcal{T}(a,s)$ the result relation for $a$ and $\mathcal{T}_{\mathcal{A}} := \bigcup_{a \in \mathcal{A}} \mathcal{T}_a$ the result relation of $\Pi$.
}{action_application}

\definition{State Space}{
The graph $\tuple{\mathcal{S}, \mathcal{T}_{\mathcal{A}}}$ is called the state space induced by \linkterm{$\Pi$}{search_problem}.
}{state_space}

\definition{Solution}{
A solution for \linkterm{$\Pi$}{search_problem} consists of a sequence $a_1, \cdots, a_n$ of actions such that for all $1 < i \leq n$:
\begin{itemize}
    \item $a_i$ is \linkterm{applicable}{action_application} to state $s_{i-1}$ where $s_o \in \mathcal{I}$, and
    \item $s_i \in \mathcal{T}_{a_i} (s_{i-1})$, and $s_n \in \mathcal{G}$
\end{itemize}
}{search_prob_sol}

\definition{Cost Function}{
Often we add a cost function $\func{c}{\mathcal{A}}{\R_0^{+}}$ that associates a step cost $c(a)$ to an action $a \in \mathcal{A}$. The cost of a \linkterm{solution}{search_prob_sol} is the sum of the step costs of its actions.
}{cost_function}

\commandnote{
\begin{itemize}
    \item For \linkterm{deterministic}{env_types} environments, we have $ \abs{\mathcal{T}(a,s)} \leq 1$
    \item For \linkterm{fully observable}{env_types} ones, we have $\mathcal{I} = \set{s_0}$ 
\end{itemize}
}

\definition{Successor Function/State}{
In a \linkterm{search problem}{search_problem}, $\mathcal{T}_a$ induces a \linkterm{partial function}{partial_function} $\pfunc{S_a}{\mathcal{S}}{\mathcal{S}}$ whose natural domain is the \linkterm{set}{def:set} of states where $a$ is \linkterm{applicable}{action_application}: $S_a(s) := s'$ if $\mathcal{T}_a = \set{s'}$ and undefined at $s$ otherwise. We call $S_a$ the \textbf{successor function} for $a$ and $S_a(s)$ the \textbf{successor state} of $s$.
}{successor_func_state}

\commandnote{
\begin{itemize}
    \item A \linkterm{search problem}{search_problem} is called a \textbf{single-state problem} iff it is \linkterm{fully observable}{env_types}, \linkterm{deterministic}{env_types}, \linkterm{static}{env_types}, and \linkterm{discrete}{env_types}.
    \item A \linkterm{search problem}{search_problem} is called a \textbf{multi-state problem} iff it is \linkterm{partially observable}{env_types}.
    \item A \linkterm{search problem}{search_problem} is called a \textbf{contingency problem} iff the environment is \linkterm{non deterministic}{env_types} and the \linkterm{state space}{state_space} is unknown.
\end{itemize}
}

\definition{Tree Search}{
Given a \linkterm{search problem}{search_problem} $\Pi := \tuple{\mathcal{S}, \mathcal{A}, \mathcal{T}, \mathcal{I}, \mathcal{G}}$, the \textbf{tree search algorithm} consists of the simulated exploration of the \linkterm{state space}{state_space} $\tuple{\mathcal{S}, \mathcal{T}_{\mathcal{A}}}$ in a search \linkterm{tree}{tree} formed by successively expanding already explored states.
}{tree_search}

\definition{Path Cost}{
We define the path cost of a node $n$ in a \linkterm{search tree}{tree_search} $T$ to be the sum of the \linkterm{step costs}{cost_function} on the \linkterm{path}{path_graph} from $n$ to the root of $T$.
}{path_cost}

The general structure of \linkterm{Tree Search}{tree_search} algorithms is as follows:
\begin{verbatim}
TREE_SEARCH(start, INSERT):
    fringe ← empty list
    append start to fringe
    while fringe is not empty:
        node ← remove first element of fringe
        if is_goal(node):
            return node
        children ← expand(node)
        INSERT(fringe, children)
    return failure
\end{verbatim}

\definition{Strategy}{
A strategy is a \linkterm{function}{function} that picks a node from the fringe of a \linkterm{search tree}{search_tee}
}{search_strategy}

\commandnote{
Every algorithm has its own \linkterm{strategy}{search_strategy}, hence the implemention of the \texttt{INSERT} differs based on that \linkterm{strategy}{search_strategy}.
}

\definition{Properties of Strategies}{
\begin{itemize}
\item completeness: does it always find a solution if one exists?
\item optimality: does it always find the optimal solution (least cost)?
\item time complexity: number of nodes the algorithm explores (expands)
\item space complexity: maximum number of nodes in memory 
\end{itemize}
}{proporties_strategies}

\commandnote{
Time and space complexity measured in terms of:
\begin{itemize}
    \item $b$: maximum branching factor of the \linkterm{search tree}{tree_search}
    \item $d$: minimal \linkterm{graph depth}{depth_graph} of a \linkterm{solution}{search_prob_sol} in the \linkterm{search tree}{tree_search}
    \item $m$: maximum \linkterm{graph depth}{depth_graph} of the \linkterm{search tree}{tree_search}
\end{itemize}
}

\definition{Breadth-First Search (BFS)}{
The BFS \linkterm{strategy}{search_strategy} treats the fringe as a FIFO queue, i.e. \linkterm{successors}{successor_func_state} go in at the end (back) of the fringe.
}{bfs}

\minititle{BFS Properties Analysis}
\begin{itemize}
\item Time Complexity:

assume we have a binary \linkterm{search tree}{tree_search} ($b=2$) and we found \linkterm{solution}{search_prob_sol} in \linkterm{depth}{depth_graph} $=3$. In the $0^{\text{th}}$ level we explored $2^0 = 1$ nodes. In the $1^{\text{st}}$ level, $2^1 = 2$ nodes. In the $2^{\text{nd}}$ level,  $2^2 = 4$ nodes. And finally $2^3 = 8$ nodes in the third level. Number of nodes the algorithm explored will be $2^0 + 2^1 + 2^2 + 2^3$.
\begin{itemize}
    \item Generally, for arbitrary $b,d$, we have $1+b+b^2 + \cdots + b^d$, which means, worst time complexity is $\mathcal{O}(b^d)$ (exponential in $d$).
\end{itemize}
\item The space complexity is also $\mathcal{O}(b^d)$.
\item \linkterm{BFS}{bfs} is complete if: (i) $b$ is finite, and (ii) the \linkterm{state space}{state_space} is finite or has a solution.
\item \linkterm{BFS}{bfs} is optimal if we have a uniform constant cost for all actions.
\end{itemize}


\definition{Uniform-Cost Search (UCS)}{
UCS is the \linkterm{strategy}{search_strategy} where the fringe is ordered by increasing \linkterm{path cost}{path_cost}. (\textit{expands least cost first})
}{ucs}

\commandnote{UCS it is equivalent to BFS if all costs are equal}

\minititle{UCS Properties Analysis}
\begin{itemize}
    \item Time and space complexity $\approx \bigo{b^d}$
    \item \linkterm{UCS}{ucs} is complete if:
    \begin{itemize}
        \item $b$ is finite
        \item all actions costs are $\geq \epsilon > 0$
        \item \linkterm{state space}{state_space} is finite or has a solution.
    \end{itemize}
\end{itemize}

\definition{Depth-First Search (DFS)}{
DFS is the \linkterm{strategy}{search_strategy} where the fringe is organized as a LIFO stack, i.e. \linkterm{successors}{successor_func_state} go in at the front of the fringe.
}{dfs}

\definition{Backtracking}{
Every node that is pushed to the stack is called a \textbf{backtrack point}. 
Popping a non-goal node from the stack and continuing the search with the new top element is called \textbf{backtracking}. 
For this reason, the \linkterm{DFS algorithm}{dfs} is also referred to as a \textbf{backtracking search}.
}{backtracking}

\minititle{DFS Properties Analysis}
\begin{itemize}
    \item Time complexity is $\bigo{b^m}$ (exponential in the maximum depth)
    \item Space complexity is linear. While \linkterm{BFS}{bfs} keeps track of all nodes at the current level in memoty, \linkterm{DFS}{dfs} needs to only keep the current path (from root to $m$), and also needs to keep the remaining unexplored sibling root node for each node along that path (to backtrack). This means, for each level, we store the siblings, i.e. $b$ nodes. We explore until we hit the max depth $m$ and in each level we store $b$. Hence we have a linear space $\bigo{b \cdot m}$
    \item \linkterm{DFS}{dfs} is complete if the \linkterm{search tree}{tree_search} is \linkterm{finite}{set_cardinality} and \linkterm{acyclic}{cyclic_graphs}.
    \item \linkterm{DFS}{dfs} is not optimal.
\end{itemize}

\definition{Depth-Limited Search (DLS)}{
DLS is a \linkterm{DFS}{dfs} with a depth limit $l$. We treat all nodes at depth $l$ as if they have no children.
}{depth_limited_search}

\minititle{DLS Properties Analysis}
\begin{itemize}
\item Time complexity $\leadsto \bigo{b^l}$
\item Space complexity $\leadsto \bigo{b \cdot l}$
\item Not complete (solution maybe beyond chosen $l$)
\item Not optimal
\end{itemize}

\definition{Iterative Deepening Search (IDS)}{
IDS is a \linkterm{DLS}{depth_limited_search} with an ever-increasing depth limit. We call the difference between successive depth limits the step size.
}{ids}

\commandnote{
\begin{itemize}
    \item \linkterm{IDS}{ids} solves the problem of choosing a good $l$ in \linkterm{DLS}{depth_limited_search}.
    \item \linkterm{IDS}{ids} tries all values until either a \linkterm{solution}{search_prob_sol} is found (depth is returned to \linkterm{DLS}{depth_limited_search}, we call that the cutoff value) or we return failure.
    \item \linkterm{IDS}{ids} combines benefits of \linkterm{DFS}{dfs} and \linkterm{BFS}{bfs}.
\end{itemize}
}

\minititle{IDS Properties Analysis}
It is similar to \linkterm{DFS}{dfs} in memory requirements (assuimg \linkterm{finite}{set_cardinality} and \linkterm{acyclic}{cyclic_graphs} \linkterm{search tree}{tree_search}).
\begin{itemize}
    \item $\bigo{b \cdot d}$ when there is a \linkterm{solution}{search_prob_sol}
    \item $\bigo{b \cdot m}$ when there is no \linkterm{solution}{search_prob_sol}
\end{itemize}

It is similar to \linkterm{BFS}{bfs} in terms of optimality, completeness, and time complexity. Optimal for problems where all actions have the same cost. Complete if $b$ is \linkterm{finite}{set_cardinality} and the \linkterm{state space}{state_space} is finite or has a solution.

With regard to time complexity, consider \linkterm{IDS}{ids} finds a solution at depth $d$, then:
\begin{itemize}
    \item nodes at the very bottom ($d$) are visited only once in the final iteration. 
    \item nodes at $d-1$ are visited twice, once in the search with depth $d-1$ and once in the search with depth $d$.
    \item nodes at $d-2$ are visited three times
    \item and so son, until root node ($d=0$) which is visited $d$ times.
\end{itemize}
Generally, we have $(d)b^1 + (d-1)b^2 + (d-2)b^3 + \cdots + b^d$. However, the majority of nodes are at the max depth where there are $b^d$ nodes. Those are only visited once and the extra cost of visiting shallower nodes does not significantly increase the overall count. Therefore, time complexity is bounded by $b^d$ ($\bigo{b^d}$) because $b^d$ dominates the total nodes count in a large \linkterm{search space}{tree_search}.


\definition{Graph Search}{
A graph search algorithm is a variant of a \linkterm{tree search algorithm}{tree_search} that prunes nodes whose state has already been considrered (duplicate pruning), essentially using a \linkterm{DAG}{dag} data strucutre.
}{graph_search}

The general structure of \linkterm{Graph Search}{graph_search} algorithms is as follows:

\begin{verbatim}
GRAPH_SEARCH(start, INSERT):
    fringe ← empty list
    visited ← empty set
    append start to fringe
    add state(start) to visited
    while fringe is not empty:
        node ← remove first element of fringe
        if is_goal(node):
            return node
        children ← expand(node)
        for each child in children:
            if state(child) not in visited:
                add state(child) to visited
                INSERT(fringe, child)
    return failure
\end{verbatim}

\definition{Search Alrogrithm}{
We speak of a \textbf{search algorithm} when we do not want to distinguish whether it is a \linkterm{tree}{tree_search} or a \linkterm{graph search algorithm}{graph_search} - \textit{difference considered as an implementation detail}.
}{search_alg}

\definition{Informed}{
A \linkterm{search algorithm}{search_alg} is called \textbf{informed}, iff it uses some form of external information to guide the search.
}{informed}

\definition{Evaluation Function}{
An evaluation function assigns a desirability value to each node of the \linkterm{search tree}{tree_search}. It is not part of the \linkterm{search problem}{search_problem}, but must be added externally.
}{eval_func}

\definition{Best-First Search}{
In best-first search, the fringe is a queue sorted in decreasing order of desirability.
}{best_first_search}

\definition{Heuristic}{
A heuristic is an \linkterm{evaluation function}{eval_func} $h$ on states that estimates the \linkterm{cost}{cost_function} form $n$ to the nearest goal state. We speak of \textbf{heuristic search} if the \linkterm{search algorithm}{search_alg} uses a heuristic in some way.
}{heuristic_search}

\definition{Greedy Search}{
A \linkterm{best-first search}{best_first_search} \linkterm{strategy}{search_strategy} where the fringe is organized as a queue sorted based on $h$ value. 
}{greedy_search}

\minititle{Greedy Search Properties}
\begin{itemize}
    \item Time and space complexity are both exponential in $m$ (max depth) $\leadsto \bigo{b^m}$
    \item It is not optimal. But it is complete if the \linkterm{state space}{state_space} is finite.
\end{itemize}

\definition{Heuristic Function}{
A heuristic funciton for \linkterm{search problem $\Pi$}{search_problem} is a \linkterm{function}{function} $\func{h}{\mathcal{S}}{\R_0^{+} \cup \{ \infty \}}$ so that $h(s) = 0 \quad \forall s \in \mathcal{G}$.
}{heuristic_function}


\definition{Goal Distance Function}{
The \linkterm{function}{function} $\func{h^{*}}{\mathcal{S}}{\R_0^{+} \cup \{ \infty \}}$, where $h^{*}(s)$ is the \linkterm{cost}{path_cost} of a cheapest \linkterm{path}{path_graph} from $s$ to a goal state, or $\infty$ if no such path exists. (\textit{the perfect heuristic which we do not know and cannot compute})
}{goal_distance_function}

\definition{Admissible}{
For a \linkterm{search problem $\Pi$}{search_problem} with states $\mathcal{S}$ and actions $\mathcal{A}$. We say that a \linkterm{heuristic}{heuristic_search} $h$ for $\Pi$ is \textbf{admissible} if
\[
h(s) \leq h^{*}(s) \quad \forall s \in \mathcal{S}
\]
}{admissible}

\commandnote{
\linkterm{Admissible}{admissible} \linkterm{heuristics}{heuristic_search} never overestimates; our guess should be always optimistic or exact, never too high 
}

\definition{Consistent}{
For a \linkterm{search problem $\Pi$}{search_problem} with states $\mathcal{S}$ and actions $\mathcal{A}$. We say that a \linkterm{heuristic}{heuristic_search} $h$ for $\Pi$ is \textbf{consistent} if 
\[
h(s) - h(s') \leq c(a) \quad \forall s \in \mathcal{S}, a \in \mathcal{A}, s' \in \mathcal{T}(s,a)
\]
}{consistent_heuristic}

\commandnote{
A \linkterm{consistent heuristic}{consistent_heuristic} never drops faster than the real cost you pay when moving through the state space. Formally, for every transition \(s \xrightarrow{a} s'\),

\[
h(s) \le c(a) + h(s')
\]

This means the heuristic respects a triangle inequality with respect to the actual \linkterm{step costs}{cost_function}. \linkterm{Consistency}{consistent_heuristic} ensures that the estimated \linkterm{total cost}{path_cost} along a \linkterm{path}{path_graph} never decreases.
}

\definition{A* Evaluation Function}{
Given a \linkterm{path cost}{path_cost} function $g$ and a 
\linkterm{heuristic}{heuristic_search} $h$, the \textbf{A* evaluation function} 
is
\[
f(n) \;=\; g(n) + h(n).
\]
It estimates the total cost of a cheapest path from the \linkterm{start state}{search_problem} 
to a goal via $n$.
}{astar_eval_function}

\definition{A* Search}{
A* search is the \linkterm{best-first search}{best_first_search} strategy that
uses the \linkterm{A* evaluation function}{astar_eval_function} $f = g + h$
to order the fringe.
}{astar_search}

\Theorem{\linkterm{A* Search}{astar_search} with \linkterm{admissible}{admissible} \linkterm{heuristic}{heuristic_search} is optimal}{thm:astar_optimal}

\definition{Dominance}{
Let $h_1$ and $h_2$ be two \linkterm{admissible}{admissible} \linkterm{heuristic}{heuristic_search}, we say that $h_2$ \textbf{dominates} $h_1$ if $h_2(s) \geq h_1(s) \quad \forall s \in \mathcal{S}$.
}{dominance}

\Theorem{If $h_2$ \linkterm{dominates}{dominance} $h_1$, then $h_2$ is better for \linkterm{search}{search_problem} than $h_1$}{thm:dominates}
\Proof{
If $h_2$ \linkterm{dominates}{dominance} $h_1$, then $h_2$ is closer to \linkterm{$h^{*}$}{goal_distance_function} than $h_1$
}

\definition{Adversarial Search}{
An adversarial search problem is a \linkterm{search problem}{search_problem} $\tuple{\mathcal{S}, \mathcal{A}, \mathcal{T}, \mathcal{I}, \mathcal{G}}$, where:
\begin{itemize}
    \item $\mathcal{S} = \mathcal{S}^{\text{Max}} \biguplus \mathcal{S}^{\text{Min}} \biguplus  \mathcal{G} $
    \item $\mathcal{A} = \mathcal{A}^{\text{Max}} \biguplus \mathcal{A}^{\text{Min}}$
    \item For $a \in \mathcal{A}^{\text{Max}}$, if $s \xrightarrow{\mathit{a}} s'$, then $s \in \mathcal{S}^{\text{Max}}$ and $s' \in (\mathcal{S}^{\text{Min}} \cup  \mathcal{G})$
    \item For $a \in \mathcal{A}^{\text{Min}}$, if $s \xrightarrow{\mathit{a}} s'$, then $s \in \mathcal{S}^{\text{Min}}$ and $s' \in (\mathcal{S}^{\text{Max}} \cup  \mathcal{G})$
\end{itemize}
together with a \textbf{game utility function} $\func{u}{\mathcal{G}}{\R}$
}{adversarial_search_problem}

\definition{Strategy}{
Let $\Theta$ be an \linkterm{adversarial search problem}{adversarial_search_problem}, and let $X \in \set{\text{Max}, \text{Min}}$. A \textbf{strategy} for $X$ is a \linkterm{function}{function} $\func{\sigma^X}{\mathcal{S}^X}{\mathcal{A}^X}$ so that $a \in \mathcal{A}^X$ is \linkterm{applicable}{action_application} to $s \in \mathcal{S}^X$ whenever $\sigma^X(s) = a$.
}{strategy_adversarial}

\definition{Optimal Strategy}{
A \linkterm{strategy}{strategy_adversarial} is called \textbf{optimal} if it yields the best possible \linkterm{utility}{adversarial_search_problem} for $X$ assuming perfect opponent play.
}{optimal_strategy}

\commandnote{
Assumptions for Adversarial Games:
\begin{itemize}
    \item Two players acting in strictly alternating turns
    \item Fully observable and deterministic game states
    \item Finite state space so the game tree is finite
    \item Terminal states have a real-valued utility
    \item The game is zero-sum: Max maximizes the utility, Min minimizes the same utility
\end{itemize}
}


\definition{Minimax}{
Let $\Theta$ be an \linkterm{adversarial search problem}{adversarial_search_problem} with \linkterm{utility function}{adversarial_search_problem} $u:\mathcal{G}\to\mathbb{R}$.  
The \textbf{minimax value} is the \linkterm{function}{function} $\hat u:\mathcal{S}\to\mathbb{R}$ defined by:

\[
\hat u(s)=
\begin{cases}
u(s) & \text{if } s\in\mathcal{G},\\[6pt]
\max\limits_{s'\in \mathcal{T}(a,s)} \hat u(s') & \text{if } s\in\mathcal{S}^{\mathrm{Max}},\\[8pt]
\min\limits_{s'\in \mathcal{T}(a,s)} \hat u(s') & \text{if } s\in\mathcal{S}^{\mathrm{Min}},
\end{cases}
\]
where $a$ ranges over all actions \linkterm{applicable}{action_application} to $s$. The \textbf{minimax decision} at a root state $r$ is any action whose \linkterm{successor state}{successor_func_state} achieves $\hat u(r)$.
}{minimax}

\definition{Minimax Algorithm}{
The minimax algorithm is given by the following recursive function whose arguments are a node (state) and whether it is Max's turn or Min's.
}{minimax_alg}

\begin{verbatim}
function MINIMAX(node, maximizing):
    if TERMINAL(node):
        return UTILITY(node)

    if maximizing:
        best = -infinity
        for each child in SUCCESSORS(node):
            val = MINIMAX(child, false)
            if val > best:
                best = val
        return best

    else:
        best = +infinity
        for each child in SUCCESSORS(node):
            val = MINIMAX(child, true)
            if val < best:
                best = val
        return best
\end{verbatim}

\definition{Evaluation Function}{
In practice, full-depth \linkterm{minimax}{minimax_alg} is infeasible because the \linkterm{search tree}{tree_search} is too large.  
We therefore impose a fixed \textit{depth limit / horizon} $d$ and stop the search at all states $\{\,s \in \mathcal{S} \mid \mathrm{dp}(s) = d\,\}$, called \textit{cut-off states}.  
An evaluation function is a \linkterm{function}{function} $f:\mathcal{S}\to\mathbb{R}$ that assigns a numerical estimate of the true \linkterm{minimax value}{minimax} $\hat u(s)$ to any nonterminal cut-off state $s$. If a cut-off state is terminal, its utility $u(s)$ is used instead of $f(s)$.
}{eval_func_minimax}


\definition{Alpha Value}{
For each node $n$ in a \linkterm{minimax}{minimax_alg} \linkterm{search tree}{tree_search}, the alpha value $\alpha(n)$ is the highest Max-node utility that search has encountered on its \linkterm{path}{path_graph} from the \linkterm{root}{tree} to $n$.
}{alpha_value}

\definition{Beta Value}{
For each node $n$ in a \linkterm{minimax}{minimax_alg} \linkterm{search tree}{tree_search}, the beta value $\beta(n)$ is the highest Min-node utility that search has encountered on its \linkterm{path}{path_graph} from the \linkterm{root}{tree} to $n$.
}{beta_value}


\definition{Alpha-Beta Pruning}{
Alpha-beta pruning is a variant of \linkterm{minimax}{minimax_alg} that avoids exploring branches that cannot change the \linkterm{minimax value}{minimax}.
\begin{itemize}
    \item At a Max-node $n$, if a successor $s$ is found with utility $\hat u(s) \ge \beta(n)$, then no remaining successors of $n$ can affect its value and they are pruned.
    \item At a Min-node $n$, if a successor $s$ is found with utility $\hat u(s) \le \alpha(n)$, then no remaining successors of $n$ can affect its value and they are pruned.
\end{itemize}
The algorithm returns the same \linkterm{minimax value}{minimax} as full search while expanding fewer nodes.
}{alpha_beta_pruning}

\begin{verbatim}
function ALPHABETA(node, alpha, beta, maximizing):
    if TERMINAL(node):
        return UTILITY(node)

    if maximizing:
        best = -infinity
        for each child in SUCCESSORS(node):
            val = ALPHABETA(child, alpha, beta, false)
            if val > best:
                best = val
            if best > alpha:
                alpha = best
            if beta <= alpha:
                break
        return best

    else:
        best = +infinity
        for each child in SUCCESSORS(node):
            val = ALPHABETA(child, alpha, beta, true)
            if val < best:
                best = val
            if best < beta:
                beta = best
            if beta <= alpha:
                break
        return best
\end{verbatim}







